### 2013-10-10 src2/thesis.tex
apa

\usepackage{multibib} 
\newcites{bb}{Bibliography} 
\newcites{ifl}{Bibliography}
\newcites{papp}{Bibliography}
\newcites{exp}{Bibliography}
\newcites{hl}{Bibliography}
\newcites{csort}{Bibliography}
\newcites{emb}{Bibliography}
\newcites{arbb}{Bibliography}
% \newcites{t}{Bibliography}
%\newcites[ifl}{Bibliography}

@@

 \textbf{Keywords:} Keyword1, Keyword2

@@

\vspace{5mm}
This research has been funded by the Swedish Foundation for
Strategic Research (which funds the Resource Aware Functional 
Programming (RAW FP) Project) and by the Swedish Research Council.

@@

Modern CPUs are parallel in a number of different ways. They discover instruction
level parallelism (ILP), they have multiple processor cores and vector units for 
Single Instruction Multiple Data parallelism. The exact configuration, number
of cores and number of vector units, vary over available processor models. 
To obtain maximum performance, with a specific target processor in mind, the programmer
would use low-level intrinsics in order to get SIMD parallelism and some threading library
for parallelism over the available cores. One problem with this is that the resulting program 
would be specialised to a specific processor model.

In chapter~\ref{chap:ArBB},

\noindent\emph{More text, lead into something about ArBB and forward ref to ArBB papers}

\noindent\emph{Data-Parallelism, Structured-Compositional programming}

Graphics Processors (GPUs), trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate much chip real estate for caches, branch prediction and 
ILP, GPUs put additional cores. This comes at a price of increased programmer effort.
GPUs have programmer managed shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application.

\noindent\emph{More text, lead into something about compositional, high-level, but 
expressive enough programming models for GPU programming and ref to Obsidian Chapter}

%\noindent\emph{GPUs}

%\subsection{Data-parallel programming on GPUs}

%\subsection{Data-parallel programming on CPUs}

\section{The problem}

\emph{what is so interesting about Parallelism and GPUs}

@@

@@

>>>>

\usepackage{multibib} 
\newcites{bb}{Bibliography} 
\newcites{ifl}{Bibliography}
\newcites{papp}{Bibliography}
\newcites{exp}{Bibliography}
\newcites{hl}{Bibliography}
\newcites{csort}{Bibliography}
\newcites{emb}{Bibliography}
\newcites{arbb}{Bibliography}
\newcites{t}{Bibliography}
%\newcites[ifl}{Bibliography}

@@

 \textbf{Keywords:} Data-parallelism, GPU, Embedded languages

@@

\vspace{5mm}
\noindent This research has been funded by the Swedish Foundation for
Strategic Research (which funds the Resource Aware Functional 
Programming (RAW FP) Project) and by the Swedish Research Council.

@@

Modern CPUs are parallel in a number of different ways. They discover instruction
level parallelism (ILP), they have multiple processor cores and vector units for 
Single Instruction Multiple Data parallelism. The exact configuration, number
of cores and number of vector units, vary over available processor models. 
To obtain maximum performance, with a specific target processor in mind, the programmer
would use low-level intrinsics in order to get SIMD parallelism and a threading library
for parallelism over the available cores. One problem with this is that the resulting program 
would be specialised to a specific processor model. The Intel Array Building Blocks (ArBB) 
system approached this problem by providing a high level programming interface~\citet{ARBB2011}
. This interface is based on a set of parallel operations on arrays, such as map, folds 
and scans. ArBB was implemented as an embedded language in C++ but a low-level C interface to 
its capabilities was also offered. ArBB programs are JIT-ed (Just-In-Time compiled) to 
target the available CPU or Accelerator (Such as the Intel Larrabee). In 
chapter~\ref{chap:ArBB}, there are papers describing work towards embedding Intel 
ArBB in Haskell.

Graphics Processors (GPUs), trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area for caches, branch prediction and 
ILP, GPUs put additional cores. This comes at a price of increased programmer effort.
GPUs have programmer managed shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application.

Intel ArBB is part of a movement towards thinking of structure and patterns in 
parallel programming. Other languages and libraries with similar aspirations 
are NVIDIA Thrust~\citet{THRUST} and Data.Array.Accelerate~\citet{ACCELERATEDAMP11}.
In the book Structured Parallel Programming by Michael McCool et al, a case is 
made for thinking in terms of buidling blocks for parallel programming~\citet{STRUCTURED}. 
They emphasize the importance of collective array operations such as map, fold, scans, 
permutations and gather/scatter operations. The book contains examples in ArBB, Cilk Plus, 
Intel Threading Building Blocks, OpenMP and OpenCL. Not all these languages include 
the patterns as abstractions, but most are possible to express in each of the languages. 
The patterns, while very powerful as abstractions, also provide tools for thinking about 
parallel algorithms.

The patterns are powerful abstractions, but it is also important not to be locked into 
a single implementation of each pattern. This could have negative performance impacts. 
In ArBB for example the patterns used are specialised to the current hardware during JIT. 
Languages like OpenCL and CUDA do not provide abstractions of these patterns. The programmer 
needs to implement them using lower level features making CUDA and OpenCL code less composable 
and harder to reuse. As an example, look at the slides from NVIDIA in reference~\citet{reduction} 
where a reduction primitive is implemented in CUDA. First a naive reduction kernel is implement 
and it is refined over many steps, taking many characteristics of the GPU into mind. What is 
is that says that we do not need to redo this exploration with each new GPU generation 
and discover a new optimal sollution?

Obsidian is our embedded language for GPU programming. We raise the level of abstraction 
compared to CUDA and provide a more composable interface. But we also want to keep 
the possible for the programmer to implement patterns in different ways. A balance 
must be struck between providing enough low-level control to get at the performance 
and still make building blocks implemented by the programmer reusable and composable. 
We want to give more compositional tools for the programmer to experiment with while 
exploring the space of various ways to implement a given pattern or primitive. 
In chapter~\ref{chap:GPUProgramming}, there are papers describing work on Obsidian.

% \section{The problem}

% \emph{what is so interesting about Parallelism and GPUs}

@@

This thesis applies embedded language methodology to data-parallel programming. An 
embedded language is implemented as a library in some host language, Haskell in this case.

@@

\bibliographystylet{alpha}
\bibliographyt{thesis}
\addcontentsline{toc}{section}{Bibliography}
